---
title: "Data Visualization with R"
author: "Armand Tossou"
date: "9/6/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Grammar of Graphics with ggplot2

Each `ggplot2` plot has 3 basic layers:
- a data layer
- an aesthetics layer
- and a geometries layer

Additional layers include:
- facets: to put multiple plots on the same canvas
- statistics: 
- coordinates
- and a theme


```{r}
#load the package
library(ggplot2)

# create a scatterplot 
pl <- ggplot(data=mtcars,aes(x=mpg,y=hp))
pl + geom_point()

```

Adding facets:


```{r}
# create a scatterplot for each cylendar type
pl <- ggplot(data=mtcars,aes(x=mpg,y=hp)) + geom_point()
pl + facet_grid(cyl ~ .)

```

Let's add a statistics layer:


```{r}
# create a scatterplot for each cylendar type, adding a smoothed line of fit
pl <- ggplot(data=mtcars,aes(x=mpg,y=hp)) + geom_point()
pl + facet_grid(cyl ~ .) + stat_smooth()

```

Next, let's add coordinates to our plot.


```{r}
# create a scatterplot for each cylendar type, adding a smoothed line of fit, and adding coordinates
pl <- ggplot(data=mtcars,aes(x=mpg,y=hp)) + geom_point()
pl2 <- pl + facet_grid(cyl ~ .) + stat_smooth()
pl2 + coord_cartesian(xlim = c(15,25))

```

Finally, let's add a theme to our plot.


```{r}

# create a scatterplot for each cylendar type, adding a smoothed line of fit, 
# and adding coordinates, and adding a theme
pl <- ggplot(data=mtcars,aes(x=mpg,y=hp)) + geom_point()
pl2 <- pl + facet_grid(cyl ~ .) + stat_smooth()
pl2 + coord_cartesian(xlim = c(15,25)) + theme_bw()

```

---
## Histograms
---

RStudio ggplot cheat sheet: 
https://www.maths.usyd.edu.au/u/UG/SM/STAT3022/r/current/Misc/data-visualization-2.1.pdf


```{r}
# install.packages("ggplot2")
library(ggplot2)

# install the dataset we'll be working with
#install.packages("ggplot2movies")
library(ggplot2movies)

# show the columns of the 'movies' dataset
colnames(movies)
```


Start out by plotting a basic histogram:

```{r}
# data & aesthetics
pl <- ggplot(movies,aes(x=rating))

# geometry
pl2 <- pl + geom_histogram(binwidth = 0.1, color='red',fill='pink',alpha=0.4)

# add labels
pl3 <- pl2 + xlab('Movie Ratings') + ylab('Count')

print(pl3)

```

There are also some advanced options that we can apply to the geometry layer.


```{r}
pl <- ggplot(movies,aes(x=rating))

pl2 <- pl + geom_histogram(binwidth = 0.1,aes(fill=..count..))

pl3 <- pl2 + xlab('Movie Ratings') + ylab('Count')

print(pl3)

```


---
## Scatterplots
---


```{r}
# we'll use the built-in 'mtcars' dataframe to plot mile per galon by car weight
df <- mtcars

# data & aesthetics
pl <- ggplot(df,aes(x=wt,y=mpg))

# geometry
print(pl + geom_point())

```

Now let's improve upon our scatterplot by playing around with:
- the size of points
- transparency


```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=wt,y=mpg))

# geometry
pl2 <- pl + geom_point(size=10,alpha=0.5)

print(pl2)

```

We can size points in our scartterplot with other features in the dataset:


```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=wt,y=mpg))

# geometry
pl2 <- pl + geom_point(size=mtcars$cyl,alpha=0.5) # sizing points based on the number of car 'cylinders'

print(pl2)

```

Pass the aesthetics parameters directly inside the geometries: 


```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=wt,y=mpg))

# geometry
pl2 <- pl + geom_point(aes(size=hp,alpha=0.5)) # sizing points based on the number of car 'horsepower'

print(pl2)

```

Let's size up points based on the number of car 'cylinders'. Note how we tell R that the 'cyl' variable is categorical (rather than continuous).


```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=wt,y=mpg))

# geometry
pl2 <- pl + geom_point(aes(size=factor(cyl),alpha=0.5))

print(pl2)

```

However, it's typically recommended to rather define shapes for categorical variables.
In the scatterplot below, we get a different geometric shape for each cylinder type.


```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=wt,y=mpg))

# geometry
pl2 <- pl + geom_point(aes(shape=factor(cyl),size=3))

print(pl2)

```

Now let's try coloring by cylinder. So we get both shapes and colors based on the levels in the 'cylinder' variable.


```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=wt,y=mpg))

# geometry
pl2 <- pl + geom_point(aes(shape=factor(cyl),color=factor(cyl),size=3))

print(pl2)
```

### Now let's play around with colors.

First, let's color all points blue.

```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=wt,y=mpg))

# geometry
pl2 <- pl + geom_point(color='blue',size=3)

print(pl2)

```


Let's define colors with HEX color values. This makes it possible to get more specific colors.

Color picker websites like this one https://www.color-hex.com/ can come in handy.


```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=wt,y=mpg))

# geometry
pl2 <- pl + geom_point(color='#43e8d8',size=3) # teal color

print(pl2)

```

Color points by gradient. Here for instance, we can color the scatterplot by 'horsepower', a continuous variable.


```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=wt,y=mpg))

# geometry
pl2 <- pl + geom_point(aes(color=hp),size=3)

# we can specify the color range
pl3 <- pl2 + scale_color_gradient(low='blue',high='red')

print(pl3)

```

---
## Barplots
---

Show a count when we're dealing with categorical data.

Let's categorize the class of vehicles vurses the count of how many times they occur in the dataset.


```{r}
# use the 'mpg' data frame that's built-in ggplot2
df <- mpg

# data & aesthetics
pl <- ggplot(df,aes(x=class))

# geometry
pl2 <- pl + geom_bar()

print(pl2)

```

Use the bar value to add a 3rd layer of data.

Let's color the bars based off the 'drive' column. This creates a stacked barplot.


```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=class))

# geometry
pl2 <- pl + geom_bar(aes(fill=drv))

print(pl2)

```

Let's adjust the position adjustment inside the barplot.

Here we create side-by-side bars instead of stacked bars.


```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=class))

# geometry
pl2 <- pl + geom_bar(aes(fill=drv),position = "dodge")

print(pl2)

```

A different type of positioning within the barplot. Let's show percentages on the bars, rather than counts of instances.


```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=class))

# geometry
pl2 <- pl + geom_bar(aes(fill=drv),position = "fill")

print(pl2)

```

---
## Boxplots
---

We graphically depict groups of data through their quartiles.

Let's create box plots of 'miles per galon' by 'cylinder' type.

```{r}
# we'll use the 'mtcars' data frame again
df <- mtcars

# data & aesthetics
pl <- ggplot(df,aes(x=factor(cyl),y=mpg))

# geometry
pl2 <- pl + geom_boxplot()

print(pl2)

```

Flip the coordinates of the boxplot by showing it vertically instead of horizontally.


```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=factor(cyl),y=mpg))

# geometry
pl2 <- pl + geom_boxplot() + coord_flip()

print(pl2)

```

Color the boxes.


```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=factor(cyl),y=mpg))

# geometry
pl2 <- pl + geom_boxplot(fill='purple')

print(pl2)

```

Color box plots based on factors of the cylinder variable.


```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=factor(cyl),y=mpg))

# geometry
pl2 <- pl + geom_boxplot(aes(fill=factor(cyl)))

print(pl2)

```

Now, let's add a theme layer.


```{r}
# data & aesthetics
pl <- ggplot(df,aes(x=factor(cyl),y=mpg))

# geometry
pl2 <- pl + geom_boxplot(aes(fill=factor(cyl))) + theme_light()

print(pl2)

```

---
## Variable Plotting
---

We'll compare two variables from the same dataset.

Let's use a '2D bin' chart to plot a heatmap of movie ratings by year.

```{r}
# use the movies dataset
library(ggplot2movies)

# data & aesthetics
pl <- ggplot(movies,aes(x=year,y=rating))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_bin2d()

print(pl2)

```


Let's change the default color map. 


```{r}
# data & aesthetics
pl <- ggplot(movies,aes(x=year,y=rating))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_bin2d()

# we can specify the color range
pl3 <- pl2 + scale_fill_gradient(low='green',high='red')

print(pl3)

```

By default, the bin sizes are just c(1,1). We can Control the bin sizes.


```{r}
# data & aesthetics
pl <- ggplot(movies,aes(x=year,y=rating))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_bin2d(binwidth=c(3,1))

# we can specify the color range
pl3 <- pl2 + scale_fill_gradient(low='blue',high='red')

print(pl3)

```
### HEX plot

We can also create a HEX plot instead of a rectangle '2d bin' chart.

Note that the `geom_hex()` option requires the `hexbin` package to be installed.


```{r}
# install the `hexbin` package
#install.packages("hexbin")

# data & aesthetics
pl <- ggplot(movies,aes(x=year,y=rating))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_hex()

# we can specify the color range
pl3 <- pl2 + scale_fill_gradient(low='blue',high='red')

print(pl3)

```

### 2D Density plot


```{r}
# data & aesthetics
pl <- ggplot(movies,aes(x=year,y=rating))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_density_2d()

print(pl2)

```

---
## Coordinates and Faceting
---

Learning to deal with coordinates will allow us to resize our plots correctly.

Faceting will allow us to place multiple plots next to one another.


### Coordinates

We'll use the 'mpg' bulit-in data set to create a scatterplot.

```{r}

# data & aesthetics
pl <- ggplot(mpg,aes(x=displ,y=hwy))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_point()

print(pl2)

```

Play around with coordinates in order to set `x` and `y` limts.


```{r}
# data & aesthetics
pl <- ggplot(mpg,aes(x=displ,y=hwy))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_point()

# set x-y limits
pl3 <- pl2 + coord_cartesian(xlim = c(1,4),ylim = c(13,30))

print(pl3)


```

Set aspect ratio. This ratio is `1:1` by default.


```{r}
# data & aesthetics
pl <- ggplot(mpg,aes(x=displ,y=hwy))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_point()

# set the 'x to y' aspect ratio to 1/3
pl3 <- pl2 + coord_fixed(ratio= 1/3) 

print(pl3)

```

### Faceting

First create a normal scatterplot.

```{r}
# data & aesthetics
pl <- ggplot(mpg,aes(x=displ,y=hwy))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_point()

print(pl2)

```

Now let's separate the normal scatterplot into facets based on the number of cylinders.

The `.` symbol in `facet_grid()` represents everything.


```{r}
# data & aesthetics
pl <- ggplot(mpg,aes(x=displ,y=hwy))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_point()

# separate the normal scatterplot into facets based on the number of cylinders
pl3 <- pl2 + facet_grid(. ~ cyl) 

print(pl3)

```

Another example: facet by drive type.


```{r}
# data & aesthetics
pl <- ggplot(mpg,aes(x=displ,y=hwy))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_point()

# separate the normal scatterplot into facets based on the number of cylinders
pl3 <- pl2 + facet_grid(drv ~ .) 

print(pl3)

```

Facet by drive type and the number of cylinders.


```{r}
# data & aesthetics
pl <- ggplot(mpg,aes(x=displ,y=hwy))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_point()

# separate the normal scatterplot into facets based on the number of cylinders
pl3 <- pl2 + facet_grid(drv ~ cyl) 

print(pl3)

```

---
## Themes
--- 

Themes allow us to quickly create beautiful looking graphs.


### Using built-in themes

Create a standard scatterplot.

```{r}
# use the built-in 'mtcars' dataset
df <- mtcars

# data & aesthetics
pl <- ggplot(mtcars,aes(x=wt,y=mpg))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_point()

print(pl2)

```

Let's apply a theme layer.

We can set a theme that applies to all our plots like this:

```{r}
theme_set(theme_minimal())

```

Now let's re-plot our basic scatter plot.

```{r}
# data & aesthetics
pl <- ggplot(mtcars,aes(x=wt,y=mpg))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_point()

print(pl2)

```

Instead, we can manually set a theme for each plot.


```{r}
# data & aesthetics
pl <- ggplot(mtcars,aes(x=wt,y=mpg))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_point()

# set theme manually
pl3 <- pl2 + theme_dark()

print(pl3)

```

### Accessing more theme options

We can install more themes by installing and calling the `ggthemes` package.


```{r}
#install.packages("ggthemes")

library(ggthemes)

```


Now we can access additional themes, like the Economist's publications.


```{r}

# data & aesthetics
pl <- ggplot(mtcars,aes(x=wt,y=mpg))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_point()

# set theme manually
pl3 <- pl2 + theme_economist()

print(pl3)

```

Imitate the style of the 'fivethirtyeight' blog.


```{r}

# data & aesthetics
pl <- ggplot(mtcars,aes(x=wt,y=mpg))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_point()

# set theme manually
pl3 <- pl2 + theme_fivethirtyeight()

print(pl3)

```

Try the Wall Street Journal theme.


```{r}

# data & aesthetics
pl <- ggplot(mtcars,aes(x=wt,y=mpg))

# geometry: check out the '2D bin' chart
pl2 <- pl + geom_point()

# set theme manually
pl3 <- pl2 + theme_wsj()

print(pl3)

```


---
## ggplot2 Exercises 
---

For the first few plots, use the mpg dataset.


```{r}
# call necessary libraries
# library(ggplot2)
# library(ggthemes)

# preview the 'mpg' dataset
head(mpg)

```


Ex1: Histogram of hwy mpg values:


```{r}
# data & aesthetics
pl <- ggplot(mpg,aes(x=hwy))

# geometry
#pl2 <- pl + geom_histogram(binwidth = 0.1, color='red',fill='pink',alpha=0.4)
pl2 <- pl + geom_histogram(fill='red',alpha=0.5)

# add labels
#pl3 <- pl2 + xlab('Highway') + ylab('Count')

print(pl2)

```

Ex2: Barplot of car counts per manufacturer with color fill defined by cyl count.


```{r}
# data & aesthetics
pl <- ggplot(mpg,aes(x=manufacturer))

# geometry
#pl2 <- pl + geom_bar(aes(fill=factor(cyl)),position = "fill")
pl2 <- pl + geom_bar(aes(fill=factor(cyl)))

# rotate labels by 90 degrees on x-asis for readability
pl3 <- pl2 + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

print(pl3)
```

Switch now to use the txhousing dataset that comes with ggplot2


```{r}

head(txhousing)

```


Ex3: Create a scatterplot of volume versus sales. Afterwards, play around with alpha and color arguments to clarify information.


```{r}
# data & aesthetics
pl <- ggplot(txhousing,aes(x=sales,y=volume))

# geometry
#pl2 <- pl + geom_point(aes(shape=factor(cyl),color=factor(cyl),size=3))
pl2 <- pl + geom_point(color="blue",alpha=0.2) + theme_gray()

print(pl2)

```


Ex4.: Add a smooth fit line to the scatterplot from above. Hint: You may need to look up `geom_smooth()`


```{r}
# data & aesthetics
pl <- ggplot(txhousing,aes(x=sales,y=volume))

# geometry
#pl2 <- pl + geom_point(aes(shape=factor(cyl),color=factor(cyl),size=3))
pl2 <- pl + geom_point(color="blue",alpha=0.2) + theme_gray()

# add a smooth fit line
#pl3 <- pl2 + geom_smooth(method = 'loess',color="red",size=1.5)
pl3 <- pl2 + geom_smooth(color="red",size=1.5,se=TRUE)

print(pl3)

```
