---
title: "Data Manipulation with R"
author: "Armand Tossou"
date: "9/5/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Guide to oUsing Dplyr

We'll install `dplyr`, call the library and explore the following built-in functions:
- filter() and slice()
- arrange()
- select() and rename()
- distinct()
- mutate() and transmute()
- summarise()
- sample_n() and sample_frac()

### Install and call the dplyr library

```{r}
# install dplyr package
#install.packages("dplyr")

library(dplyr)

# install some data packages
## a large data set of flights from 2013
#install.packages("nycflights13")
library(nycflights13)

# call the flights data
head(flights)

summary(flights)

```


### Explore the filter() function

`filter()` allows us to select as subset of rows in a dataframe.

```{r}
# flights from American Airlines that occured on November 3rd.
head(filter(flights, month==11, day==3, carrier=='AA'))

# trying bracket notation to achieve the same filter result
head(flights[flights$month==11 & flights$day==3 & flights$carrier=='AA',])

```


### Explore the slice() function 

`slice()` allows us to select rows by position.

```{r}
# first 10 rows
slice(flights, 1:10)

```


### Explore the arrange() function

`arrange()` works similarly to filter(), except it re-orders the rows instead of selecting/filtering them.

```{r}
# first 6 rows, ordered by some key columns
head(arrange(flights,year,month,day,arr_time))

# first 6 rows, ordered by some key columns
## do descending order for arrival time
head(arrange(flights,year,month,day,desc(arr_time)))

```

### Explore the select() function

`select()` allows us to zoom in on a few columns of interest.


```{r}

head(select(flights, carrier, arr_time, month))


```

### Explore the rename() function

`rename()` allows us to quickly rename column names. 


```{r}

# let's rename the 'carrier' column as 'airline_carrier'
head(rename(flights, airline_carrier = carrier))

```

### Explore the distinct() function

`distinct()` allows us to select the unique values in a column.


```{r}
# find out the list of unique aireline carriers in the flights dataset
distinct(select(flights,carrier))

```

### Explore the mutate() function

`mutate()` allows us to add new columns that are functions of existing columns.


```{r}
# create a new column that is the difference between arrival delay and departure delay
mutate(flights,new_col = arr_delay - dep_delay)

```

### Explore the transmute() function

`transmute()` is related to `mutate()`. We use it if we just want the newly created column back.


```{r}
# create a new column that is the difference between arrival delay and departure delay
transmute(flights,new_col = arr_delay - dep_delay)

```

### Explore the summarise() function

`summarise()` allows us to collapse the dataframe into single rows using some sort of a function that aggregate results.   


```{r}

# compute average air time, with missing values removed
summarise(flights, avg_air_time=mean(air_time,na.rm=TRUE))

# compute total airtime for all the flights
summarise(flights,total_time=sum(air_time,na.rm=TRUE))

```

### Explore the sample_n() function

`sample_n()` allows us to take random samples of the data frame, by specifying a number to pick.


```{r}

# take a random sample of 10 rows from the flights data frame
sample_n(flights,10)

```

### Explore the sample_frac() function

`sample_frac()` allows us to take random samples of the data frame, by specifying a percentage to pick. 


```{r}

# take a random sample of 10% of the rows from the flights data frame
sample_frac(flights,0.1)

```

---
## Pipe operator
---

The pipe operator `%>%` is really handy while working with `dplyr` libraries. It allows us to chain multiple operations together.


```{r}
#load the built-in mtcars data frame
df <- mtcars
  
# motivation for the pipe %>% operator:
## nesting: makes interpretation of the code hard
result <- arrange(sample_n(filter(df,mpg>20), size=5),desc(mpg))

print(result)

# using a multiple assignments approach instead
a <- filter(df,mpg>20)
b <- sample_n(a, size = 5)
result <- arrange(b,desc(mpg))

print(result)

# re-write all of this using the pipe operator: %>%
## syntax: Data %>% operation1 %>% operation2 %>% operation3 ...
results <- df %>% filter(mpg>20) %>% sample_n(size = 5) %>% arrange(desc(mpg))
print(result)

```


---
## Dplyr Exercises 
---

We will use the mtcars dataframe for this exercise!

```{r}

head(mtcars)

```

### EX1. Return rows of cars that have an mpg value greater than 20 and 6 cylinders.


```{r}

filter(mtcars,mpg>20,cyl==6)

```

### Ex2. Reorder the Data Frame by cyl first, then by descending wt.


```{r}

arrange(mtcars,cyl,desc(wt))

```

### Ex3. Select the columns mpg and hp.


```{r}

select(mtcars,mpg,hp)

```

### Ex4. Select the distinct values of the gear column.


```{r}

distinct(mtcars,gear)


```

### Ex5. Create a new column called "Performance" which is calculated by hp divided by wt.


```{r}

mutate(mtcars,performance = hp / wt)

```

### Ex6. Find the mean mpg value using dplyr.


```{r}

summarise(mtcars,avg_mpg=mean(mpg))


```

### Ex7. Use pipe operators to get the mean hp value for cars with 6 cylinders.


```{r}

mtcars %>% filter(cyl==6) %>% summarise(avg_hp=mean(hp,na.rm=TRUE))

```


---
## Guide to using Tidyr
---

- Install and load packages
- Explore the four most important functions in the `tidyr` library: 
  - `gather`
  - `spread`
  - `separate`
  - and `unite`
- Explore the `data.table` package. A `data.table` is essentially a `data.frame` with a few more features. This offers the advantages of speed and a cleaner syntax.

```{r}
# install the tidyr package
#install.packages("tidyr")
library(tidyr)

# install a complementary data package

#install.packages("data.table")
library(data.table)

```

### Explore the `gather` function

Collapse multiple columns into key-pair values.


```{r}
# let's create a data frame of quarterly returns for a company
comp  <- c(1,1,1,2,2,2,3,3,3)
yr <- c(1998:2000)
q1 <- runif(9, min = 0, max = 100)
q2 <- runif(9, min = 0, max = 100)
q3 <- runif(9, min = 0, max = 100)
q4 <- runif(9, min = 0, max = 100)

df <- data.frame(comp=comp,year=yr,Qtr1=q1,Qtr2=q2,Qtr3=q3,Qtr4=q4)
df

```

This data frame is in wide format.

```{r}
# Use the gather() function
# to restructure the time component of the data into long format
df_long <- gather(df,Quarter,Revenue,Qtr1:Qtr4)
df_long

```

### Explore the `spread` function

`spread()` is complementary to `gather()`.


```{r}
# create a stocks dataset
stocks <- data.frame(
  time = as.Date('2009-01-01') + 0:9,
  x = rnorm(10, 0, 1),
  y = rnorm(10, 0, 2),
  z = rnorm(10, 0, 4)
)

stocks

# gather the data
stocks.gathered <- stocks %>% gather(stock,price,x,y,z)
head(stocks.gathered)

# let's spread this dataset back out
## this will generate a data frame in wide format
stocks.gathered %>% spread(stock,price)

# another way to spread the data
spread(stocks.gathered,time,price)

```

### Explore the separate function

`separate` allows us to convert a single character column into multiple columns, using regular expressions or vector / character expressions.


```{r}
# let's generate a dataframe
df <- data.frame(new.col=c(NA,"a.x","b.y","c.z"))
df

separate(df,new.col,c("ABC","XYZ"))

```

The `separate()` function uses a non-alphanumeric character to split the column name. Now let's specify the separator.

```{r}
# let's generate a dataframe
df <- data.frame(new.col=c(NA,"a-x","b-y","c-z"))
df

df.sep <- separate(data=df,col=new.col,into=c("ABC","XYZ"),sep = "-")
df.sep

```

### Explore the unite function

`unite()` is the exact opposite of `separate()`.

```{r}

unite(data=df.sep,col=new.joined.col,ABC,XYZ,sep = "---")



```
