---
title: "Interactive Visualizations with Plotly"
author: "Armand Tossou"
date: "9/6/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# add this line to avoid errors in compilation. See: https://stackoverflow.com/questions/30519673/cant-resolve-error-in-rmd-file-anonymous-withcallinghandlers-withvisi
knitr::opts_chunk$set(error = TRUE)
```

## Some resources

Resources and Links for Plotly and ggplot2.


Here are the links for the resources mentioned in this lecture:

[Plotly:](https://plot.ly/)

https://plot.ly


Package to install:

`install.packages('plotly')`


[Reference for gglotly() examples:](https://plot.ly/ggplot2/)

https://plot.ly/ggplot2/


[Reference for full ggplotly example documentation:](http://ropensci.github.io/plotly-test-table/tables/0e3d5ca144d27d8416318824c1b6ec1421a51045/index.html)

http://ropensci.github.io/plotly-test-table/tables/0e3d5ca144d27d8416318824c1b6ec1421a51045/index.html


## Background

We can use the open source library called 'plotly' to convert our ggplot2 graphics into interactive images.

Let's install the 'plotly' library and call all necessary libraries for our project:

```{r}
# install.packages("plotly")

# call the libraries
library(ggplot2)
library(plotly)

```


## Create a ggplot2 plot

Let's create a scatterplot from the 'mtcars' built-in dataset.

```{r}
pl <- ggplot(mtcars,aes(mpg,wt)) + geom_point()
pl

```


## Create an interactive plot

Use `ggplotly()` to convert the non-interactive plot into an interactive one.


```{r}
gpl <- ggplotly(pl)
gpl
```

## Refer to the online documentation on 'ggplotly' for sample codes and examples of various statistical charts.
