---
title: "R data frame"
author: "Armand Tossou"
date: "9/3/2021"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Sample public data set (data frame), built into R

-   The `state.x77` data set.
-   The `USPersonalExpenditure` data set.
-   The `women` data set.
-   The `WorldPhones` data set.

```{r }
state.x77

USPersonalExpenditure

women

WorldPhones

data() # list of all data sets built in R

# preview the first 6 rows of the 'state.x77' data set
head(state.x77)


# preview the last 6 rows of the 'state.x77' data set
tail(state.x77)

# get the structure of the 'state.x77' data set
str(state.x77)

# summary statistics of variables in the 'state.x77' data set
summary(state.x77)

```

## Create our own data frame

We'll create weather data.

```{r }
days <- c('Mon','Tue','Wed','Thu','Fri') # weekdays
temp <- c(22.2,21,23,24.3,25) # daily temperature
rain <- c(T,T,F,F,T) # whether or not it rained on a given day

# combine those matrices into a data frame
df <- data.frame(days,temp,rain)
df

# check the structure of the newly created data frame
str(df)

#get summary of each variable
summary(df)
```

## Selecting and Indexing Data Frame Elements

```{r }
# select everything from the first row
df[1,]

# select everything from the first column
df[,1]

## We can also index using column and row labels

# select everything from the 'rain' column
df[,'rain']
class(df[,'days']) # character

# select first 5 rows in columns 'days' and 'temp'
df[1:5,c('days','temp')]

## grab all the values of a particular column

# using the '$' notation 
df$days # R returns a vector
class(df$'days') # data frame

# using bracket notation 
df['days'] # R returns a data frame
class(df['days']) # character

## using the `subset()` function for selection

# find out days where it rained 
subset(df, subset = rain==T) # one option
subset(df,df$rain==T) # an alternative

# grab days where the temperature was greater than 23 degrees
subset(df, subset = temp > 23)

## sorting a data frame. Use the `order()` function

# sorting daily temperatures in accenting order
sorted.temp <- order(df['temp']) # this is a mask containing row indices
sorted.temp

df[sorted.temp,] # apply mask. We sort the data frame in 

# Now let's sort the data frame in descending order
desc.temp <- order(-df['temp'])
df[desc.temp,] # apply mask

```

---
## Overview of Data Frame Operations
---

Here we'll cover the following: - Creating data frames - Importing and
exporting data - Getting information about a data frame - Referencing
cells - Referencing rows - Referencing columns - Adding rows - Adding
columns - Setting column names - Selecting multiple columns - Dealing
with missing data

### Creating data frames

```{r }

# create an empty dataframe 
empty <- data.frame()

# create a dataframe from vectors

c1 <- 1:10 # numbers, 1 through 10
letters # the letters of the alphabet. They're built into R
c2 <- letters[1:10] # create a vector of the first 10 letters of the alphabet
c2

# combine c1 and c2 into a dataframe

df <- data.frame(c1,c2) # without column names assigned
df

df <- data.frame(col.name.1=c1,col.name.2=c2) # assign each column a name
df


```

### Importing and exporting data

```{r }

# to write a .csv file 
## this saves the last dataframe (i.e., df) to our working directory
## this also saves our index column as a separate column
write.csv(df, file = 'saved_df.csv')

#  reading a .csv file into R
df2 <- read.csv('saved_df.csv')
df2

```

### Getting information about a data frame

```{r }
# find out the numbers of rows and columns
nrow(df)
ncol(df)

# find out the names of rows and columns
colnames(df)
rownames(df) # these are the indices too

# get the structure of the dataframe
str(df)

# get a summary of the dataframe
summary(df)

```

### Referencing cells

```{r }
# reference a cell in the dataframe, in row 5 and column 2
df[[5,2]] # just using row and column indices
df[[5,'col.name.2']] # by using the column name

# change a value in the dataframe
df[[2,'col.name.1']] <- 9999 # change value 2 at the intersection of row 2 and column 1
df

```

### Referencing rows

```{r }
df[1,] # show row 1. Returns a dataframe
class(df[1,]) # "data.frame"

as.numeric(df[1,]) # show row 1. Returns a vector instead of a dataframe
class(as.numeric(df[1,])) # "numeric"

```

### Referencing columns

```{r }
head(mtcars) # check out this built-in dataframe

# 4 ways to grab the 'mpg' column as a vector
mtcars$mpg

mtcars[,'mpg']

mtcars[,1] # 'mpg' is the first column

mtcars[['mpg']]

# 2 ways to grab the 'mpg' column as a dataframe
mtcars['mpg']

mtcars[1] # same result using the index/order of the column

## selecting multiple columns as a dataframe
head(mtcars[c('mpg','disp')])

```

### Adding rows

```{r }
# create a new dataframe
df2 <- data.frame(col.name.1 = 2000, col.name.2 = 'new')
df2

# combine this last dataframe into an existing one
dfnew <- rbind(df,df2)
dfnew[11,] # preview the 11th row in the dataframe

```

### Adding columns

```{r }
# add a new column to 'df' that's twice the values in column 'col.name.1'
df$newcol <- 2*df$col.name.1
df

# alternative approach
df['newcol.copy'] <- df$newcol
df

```

### Setting column names

```{r }
# retrive column names of the 'df' dataframe
colnames(df)

# rename the columns
colnames(df) <- c('1','2','3','4')
df

# rename a specific column in the dataframe
colnames(df)[1] <- c('NEW COL NAME')
df

```

### Selecting multiple ROWS

```{r }
# SELECT FIRST 10 ROWS
df[1:10,]

# using the head() method, select first 7 rows
head(df,7)

# select everything but row 2
df[-2,]

## conditional selecting
head(mtcars) # use the 'mtcars' built-in dataset

# select on rows where 'mpg' is greater than 20
mtcars[mtcars$mpg > 20, ]

# using multiple conditions
mtcars[ (mtcars$mpg>20 & mtcars$cyl==6), ]
mtcars[ (mtcars$mpg>20 & mtcars$cyl==6), c('mpg','cyl','hp')]

## using the subset() function
subset(mtcars, (mpg>20 & cyl==6) )

```

### Selecting multiple columns

```{r }
# select multiple columns from 'mtcars'
mtcars[,c(1,2,3)] # first 3 columns
mtcars[,c('mpg','cyl','disp')]

```

### Dealing with missing data

```{r }
sum(is.na(df)) # get booleans

any(is.na(df)) # get a single FALSE/TRUE boolean for whether we have at leat 1 case of missing values

sum(is.na(df)) # get total number of missing values

## replace all NULL values with zero
df[is.na(df)] <- 0

## replace missing values in column 'mpg' of built-in dataset 'mtcars' by the column mean
mtcars$mpg[is.na(mtcars$mpg)] <- mean(mtcars$mpg)

```

## 

```{r }



```

---
## Data Frame Training Exercise
---

Ex 1: Recreate the following dataframe by creating vectors and using the
data.frame function:

```{r }
Ages <- c(22,25,26)
Weight <- c(150,165,120)
Sex <- c('M','M','F')

mydf <- data.frame(Ages, Weight, Sex, row.names = c('Sam','Frank','Amy'))
mydf

```

Ex 2: Check if mtcars is a dataframe using is.data.frame()

```{r }
is.data.frame(mtcars)

```

Ex 3: Use as.data.frame() to convert a matrix into a dataframe:

```{r }
mat <- matrix(1:25,nrow = 5)
df_mat <- as.data.frame(mat)
df_mat

```

Ex 4: Set the built-in data frame mtcars as a variable df. We'll use
this df variable for the rest of the exercises.

```{r }

df <- mtcars
df

```

Ex 5: Display the first 6 rows of df

```{r }
head(df,6)

# df[1:6,] # alternatively

```

Ex 6: What is the average mpg value for all the cars?

```{r }

mean(df$mpg)

```

Ex 7: Select the rows where all cars have 6 cylinders (cyl column)

```{r }

df[df$cyl == 6,]

```

Ex 8: Select the columns am,gear, and carb.

```{r }

df[ , c('am','gear','carb')]

# subset(df, select = c(am, gear, carb)) # alternative approach
```

Ex 9: Create a new column called performance, which is calculated by
hp/wt.

```{r }

df$performance <- df$hp / df$wt
head(df)

```

Ex 10: Your performance column will have several decimal place
precision. Figure out how to use round() (check help(round)) to reduce
this accuracy to only 2 decimal places.

```{r }

df$performance <- round(df$hp / df$wt, 2)
head(df)

```

Ex 11: What is the mpg of the Hornet Sportabout?

```{r }

df[['Hornet Sportabout', 'mpg']]

```
