---
title: "R Matrices"
author: "Armand Tossou"
date: "9/3/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Creating a Matrix


```{r }

# sequential numeric vector
1:10

v <- 1:10
v

matrix(v)

matrix(v, nrow = 2) # create a 2x5 matrix

matrix(1:12,byrow = FALSE, nrow = 4)

```

## Google vs Microsoft stock prices

combine the vectors:

```{r }
goog <- c(450,451,452,445,468)
msft <- c(230,231,232,233,220)

stocks <- c(goog,msft) # combine the 2 vectors

# build a matrix out of this
stock.matrix <- matrix(stocks, byrow = TRUE, nrow = 2)

days <- c('Mon','Tue','Wed','Thu','Fri')
st.names <- c('GOOG','MSFT')

colnames(stock.matrix) <- days
rownames(stock.matrix) <- st.names

print(stock.matrix)

```

## Matrix Arithmetic

```{r }

mat <- matrix(1:25, byrow=TRUE, nrow=5)
mat
print('--------------------------------')

mat * 2 # multiply everything by 2
print('--------------------------------')

mat / 2 # divide everything by 2
print('--------------------------------')

mat ^ 2 # everything to the power of 2
print('--------------------------------')

1 / mat # reciprocal of everything
print('--------------------------------')

mat > 15 # everywhere the matrix is greater than 15
mat[mat > 15]
print('--------------------------------')

mat * mat # element by element multiplication
print('--------------------------------')

mat %*% mat # matrix multiplication
```

## Matrix operations

```{r }

colSums(stock.matrix)

rowSums(stock.matrix)

rowMeans(stock.matrix)

colMeans(stock.matrix)

```

# adding rows and columns


```{r }
# add new row to matrix
FB <- c(111,112,113,120,145)
tech.stocks <- rbind(stock.matrix, FB)
tech.stocks

# add new columns to matrix
print("-------------------------------")
avg <- rowMeans(tech.stocks)
tech.stocks <- cbind(tech.stocks,avg)
tech.stocks

```

## Matrix Selection and Indexing


```{r }
mat <- matrix(1:50, byrow = TRUE, nrow = 5)
mat

mat[1,] # select 1st row

mat[,1] # select 1st column

mat[1:3,] # first 3 rows

mat[1:2,1:3] # top left rectangle

mat[,9:10] # last 2 columns

mat[2:3,5:6]# center square

```

## Factor and Categorical Matrices


```{r }
animal <- c('d','c','d','c','c')
id <- c(1:5)

factor(animal)

fact.ani <- factor(animal)

# assign orders, for ordinal matrix
ord.cat <- c('cold','med','hot')
temps <- c('cold','med','hot','hot','hot','cold','med')
fact.temps <- factor(temps,ordered = TRUE, levels=c('cold','med','hot'))
fact.temps

summary(fact.temps)
summary(temps)

```

## Exercise



Ex 1: Create 2 vectors A and B, where A is (1,2,3) and B is (4,5,6). With these vectors, use the cbind() or rbind() function to create a 2 by 3 matrix from the vectors. You'll need to figure out which of these binding functions is the correct choice.

```{r }
A <- c(1,2,3)
B <- c(4,5,6)

C <- rbind(A,B)
C
```

Ex 2: Create a 3 by 3 matrix consisting of the numbers 1-9. Create this matrix using the shortcut 1:9 and by specifying the nrow argument in the matrix() function call. Assign this matrix to the variable mat

```{r }
mat <- matrix(1:9, byrow = FALSE, nrow = 3)

mat
```

Ex 3: Confirm that mat is a matrix using is.matrix()

```{r }
is.matrix(mat)

```

Ex 4: Create a 5 by 5 matrix consisting of the numbers 1-25 and assign it to the variable mat2. The top row should be the numbers 1-5.

```{r }
mat2 <- matrix(1:25, byrow = TRUE, nrow = 5)

mat2

```

Ex 5: Using indexing notation, grab a sub-section of mat2 from the previous exercise that looks like this:

```{r }
mat2[2:3,2:3]

```

Ex 6: Using indexing notation, grab a sub-section of mat2 from the previous exercise that looks like this:

```{r }

mat2[4:5,4:5]


```

Ex 7: What is the sum of all the elements in mat2?


```{r }

sum(mat2)

```

Ex 8: Ok time for our last exercise! Find out how to use runif() to create a 4 by 5 matrix consisting of 20 random numbers (4*5=20).


```{r }

mat3 <- matrix(runif(20,0,1),nrow = 4)
mat3

```


Note: Install this library to enable document knitting into PDF:

```{r}
tinytex::install_tinytex()
```
