---
title: "R Programming Basics"
author: "Armand Tossou"
date: "9/5/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Logical Operators

- & # AND
- | # OR
- ! # NOT


```{r}
x <- 10

(x>5) & (x < 20) # TRUE

(10 == 1) # FALSE
!(10 == 1) # TRUE

## a more realistic example
df <- mtcars
head(mtcars)

# what cars have miles per galon greater than 20?
df[df$mpg > 20, ]

subset(df, mpg > 20) # alternatively

# using 2 conditions
df[(df$mpg >20) & (df$hp > 100),]

```

## If, else, and else if Statements



```{r}

# assign value to x
x <- 13

# create conditional statements
if (x==10) {
  # code
  print('x is equal to 10!')
} else if (x==12) {
  print('x is equal to 12')
} else {
  print('x was not equal to 10 or 12')
}
  

```
Another example:

```{r}
temp <- 75

if (temp > 80) {# Execute if condition was TRUE
  print('Hot outside!')
} else if (temp < 80 & temp >= 50) {
    print('Nice outside!')
} else {
    print('It\'s less than 50 degrees outside')
  }

```

## Logical Operators with Vectors

We have two options when use logical operators, a comparison of the entire vectors element by element, or just a comparison of the first elements in the vectors, to make sure the output is a single Logical.

``` {r}
tf <- c(TRUE,FALSE)
tt <- c(TRUE,TRUE)
ft <- c(FALSE, TRUE)

# create the condition
tt & tf # TRUE FALSE

tt | tf # TRUE TRUE

```


To compare first elements use && or ||.

``` {r}
ft && tt # FALSE

tt && tf # TRUE

tt || tf # TRUE

tt || ft # TRUE

```


---

## Conditional Statements Exercises

---

Example: Write a script that prints "Hello" if the variable x is equal to 1:

``` {r}
x <- 1 # Change x to test

if (x==1) {
  print('Hello')
}
```

Ex 1: Write a script that will print "Even Number" if the variable x is an even number, otherwise print "Not Even":

``` {r}
x <- 3 # Change x to test

if (x %% 2 == 0) {
  print("Even number")
} else {
  print("Not even")
}

```

Ex 2: Write a script that will print 'Is a Matrix' if the variable x is a matrix, otherwise print "Not a Matrix". Hint: You may want to check out help(is.matrix)


```{r}
x <- matrix()

if (is.matrix(x)) {
  print("Is a Matrix")
} else {
  print("Not a Matrix")
}

```

Ex 3: Create a script that given a numeric vector x with a length 3, will print out the elements in order from high to low. You must use if,else if, and else statements for your logic. (This code will be relatively long)


```{r}
#x <- c(3,7,1) # change order to test
#x <- c(1,3,7) # change order to test
x <- c(1,7,3) # change order to test

# print highest number
if ( x[1] == max(x) ) { 
  print(x[1])
} else if ( x[2] == max(x) ) {
  print(x[2])
} else {
  print(x[3])
} 

# print middle number
if ( (x[1] > x[2] & x[1] < x[3]) | (x[1] > x[3] & x[1] < x[2]) ) {
  print(x[1])
} else if ( (x[2] > x[1] & x[2] < x[3]) | (x[2] > x[3] & x[2] < x[1]) ) {
  print(x[2])
} else {
  print(x[3])
} 

# print lowest number
if ( (x[1] == min(x)) ) { 
  print(x[1])
} else if ( (x[2] == min(x)) ) {
  print(x[2])
} else {
  print(x[3])
}

```

Ex 4: Write a script that uses if,else if, and else statements to print the max element in a numeric vector with 3 elements.


```{r}
x <- c(20, 10, 1) # change order to test

# print highest number
if ( x[1] == max(x) ) { 
  print(x[1])
} else if ( x[2] == max(x) ) {
  print(x[2])
} else {
  print(x[3])
} 

```

---
## While Loops
---


```{r}
x <- 0

while (x<10) {
  print( paste0('x is: ' , x) )
  x <- x+1
  if (x==10) {
    print("x is now equal to 10! Break loop!")
    break
    print("Wooo I printed too!")
  }
}


```


```{r}
x <- 0

while (x<10) {
  print( paste0('x is: ' , x) )
  x <- x+1
  if (x==5) {
    print("x is now equal to 5! Break loop!")
    break
  }
}

```

## For Loops

A 'for loop' allows us to iterate over objects.

```{r}
v <- c(1,2,3)

for (variable in v) {
  print(variable)
}

```


```{r}
v <- c(1,2,3,4,5)

for (temp.var in v) {
  # Execute some code
  # for every temp.var in v
  print('Hello')
}

```


```{r}
v <- c(1,2,3,4,5)

for (temp.var in v) {
  result <- temp.var + 1
  print( paste0('The temp.var plus 1 is equal to:', result) )
  print('----------------------------------')
}

```

We can use for loops for other objects as well. Here's an example to illustrate the case of a list


```{r}
my.list <- list(c(1,2,3), mtcars)

for (item in my.list) {
  print(item)
}

```

Using a for loop with a matrix:


```{r}
mat <- matrix(1:25, byrow = TRUE, nrow = 5)

for (num in mat) {
  print(num)
}

```

Nested for loops:


```{r}
mat <- matrix(1:25, byrow = FALSE, nrow = 5)
mat
print("")

for (row in 1:nrow(mat)) {
  for (col in 1:ncol(mat)) {
    print( paste('The element at row:', row, 'and col:', col, 'is', mat[row,col]))
  }
}

```

---
## Functions
--

A function is just a useful device that groups together a set of statements so they can be run more than once.


```{r}
# syntax

name_of_func <- function(input1,input2,input3=45) {
  # code to execute
  result <- input1 + input2 * input3
  return(result)
}

```


Built-in functions:


```{r}
sum(c(1,2,3))
help(sum)

```

Create a simple function that takes no input:


```{r}
hello <- function() {
  # code that executes when the function is called
  print("Hello")
}

hello() # call the function

```

Create a function that takes an input:


```{r}
hello <- function(name) {
  print( paste("Hello", name))
}

hello('Sammy') # call the function

```

Assign a default value to the input:


```{r}
hello <- function(name='Frank') {
  print( paste("Hello", name))
}

hello() # call the function
hello('Sammy') # call the function

```

Create a function with multiple inputs:


```{r}
add_num <- function(num1,num2) {
  print(num1+num2)
}

add_num(4,5)
```

Examples showing how we can return a function output:


```{r}
add_num <- function(num1,num2) {
  my.sum <- num1+num2
  return(my.sum)
}

result <- add_num(4,5)
result
```

Variable scope:


```{r}
times5 <- function(num){
  return(num*5)
}

print(times5(20))
```

Same function, with some additions:


```{r}
times5 <- function(num){
  my.result <- num*5
  return(my.result)
}

my.output <- times5(100)

print(my.output)

#my.result # we get this error message 'Error: object 'my.result' not found' because the variable is local to the function

```

Showing an example of a global vs local variable:


```{r}
v <- "I'm a global variable"
stuff <- "I'm a global stuff"

fun <- function(stuff){
  print(v)
  stuff <- "Reassign stuff inside of this function fun"
  print(stuff)
} 

fun(stuff) # call the function

print(stuff)
```

---
## Function Exercise
---

EXAMPLE 1: Create a function that takes in a name as a string argument, and prints out "Hello name"


```{r}
hello <- function(name){
  print(paste('Hello', name))
}

hello('Armand')

```

EXAMPLE 2: Create a function that takes in a name as a string argument and returns a string of the form - "Hello name"


```{r}
hello_2 <- function(name){
  result <- paste('Hello', name, sep=" ")
}

print(hello_2('Armand'))

```

Ex 1: Create a function that will return the product of two integers.


```{r}
prod <- function(num1,num2) {
  return(num1 * num2)
}

prod(3,4)

```

Ex 2: Create a function that accepts two arguments, an integer and a vector of integers. It returns TRUE if the integer is present in the vector, otherwise it returns FALSE. Make sure you pay careful attention to your placement of the return(FALSE) line in your function!


```{r}
num_check <- function(int,vec) {
  if (int %in% vec) {
    return(TRUE)
  } else {
    return(FALSE)
  }
}

#num_check(2,c(1,2,3)) # change to test
num_check(7,c(1,2,3)) # change to test

```

Ex 3: Create a function that accepts two arguments, an integer and a vector of integers. It returns the count of the number of occurrences of the integer in the input vector.


```{r}
num_count <- function(num,vec){
  # create a counter
  myCounter = 0
  for(i in vec){
    if(num == i){
      myCounter = myCounter+1
    }
  }
  return(myCounter)
}

#num_count(2,c(1,1,2,2,3,3)) # test numero 1. Answer: 2
num_count(1,c(1,1,2,2,3,1,4,5,5,2,2,1,3)) # test numero 2. Answer: 4
```

Ex 4: We want to ship bars of aluminum. We will create a function that accepts an integer representing the requested kilograms of aluminum for the package to be shipped. To fullfill these orders, we have small bars (1 kilogram each) and big bars (5 kilograms each). Return the least number of bars needed.

For example, a load of 6 kg requires a minimum of two bars (1 5kg bars and 1 1kg bars). A load of 17 kg requires a minimum of 5 bars (3 5kg bars and 2 1kg bars).


```{r}
minNumBars <- function(num){
  return( (num %/% 5) + (num %% 5) )
}

minNumBars(17) # Answer: 5
minNumBars(6) # Answer: 2

```

Ex 5: Create a function that accepts 3 integer values and returns their sum. However, if an integer value is evenly divisible by 3, then it does not count towards the sum. Return zero if all numbers are evenly divisible by 3. Hint: You may want to use the append() function.

```{r}
summer <- function(int1,int2,int3){
  # create empty list
  ans <- 0
  # create conditions
  for (i in c(int1,int2,int3)){
    if (!(i %% 3==0)) {
      ans <- ans + i
    }
  } 
  return(ans)
}

#summer(7,2,3) # Answer: 9
#summer(3,6,9) # Answer: 0 
summer(9,11,12) # Answer: 11

```

Alternative solution:

```{r}
summer <- function(a,b,c){
  # create output list
  out <- c(0)
  # create conditions
  if ( a %% 3 != 0){
    out <- append(out,a)
  } 
  if ( b %% 3 != 0){
    out <- append(out,b)
  } 
  if ( c %% 3 != 0){
    out <- append(out,c)
  } 
  return(sum(out))
}

#summer(7,2,3) # Answer: 9
#summer(3,6,9) # Answer: 0 
summer(9,11,12) # Answer: 11

```
Ex 6: Create a function that will return TRUE if an input integer is prime. Otherwise, return FALSE. You may want to look into the any() function. There are many possible solutions to this problem.


```{r}
prime_check <- function(num){
  if (num<=1){
    return(FALSE)
  }
  if (num==2){
    return(TRUE)
  }
  for (x in 2:(num-1)){
    if (num %% x ==0){
      return(FALSE)
    }
  }
  return(TRUE)
}

prime_check(-30)

```

